---
title: "Project1"
author: "Mark Shocklee"
date: "Saturday, November 08, 2014"
output: html_document
---

## Setup

```{r load_packages, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
#install.packages("ggplot2", dependencies = TRUE)
library(ggplot2)
```

## Loading and preprocessing the data

Show any code that is needed to:

* Load the data (i.e. read.csv())
* Process/transform the data (if necessary) into a format suitable for your analysis

The file for this project was downloaded on 11/8/2014 10:17 AM from http://github.com/rdpeng/RepData_PeerAssessment1.

```{r load_and_preprocess, echo = TRUE, warning=FALSE}
dir.create("data")
unzip("activity.zip", exdir = "./data", overwrite=TRUE)
activity <- read.csv("./data/activity.csv", na.strings = "NA")
dir.create("figure")
```

Get some basic information about the file.

```{r basic_data_info, echo = TRUE}
summary(activity)
head(activity)
tail(activity)
```

## Processing prior to Imputing Missing Values

* What is mean total number of steps taken per day?
```{r total_steps_per_day, echo = TRUE}
total_steps <- aggregate(activity$steps, list(Date = activity$date), sum)
total_steps
mean(total_steps$x, na.rm = TRUE)
median(total_steps$x, na.rm = TRUE)
```

For this part of the assignment, you can ignore the missing values in the dataset.

* Make a histogram of the total number of steps taken each day

```{r total_steps_histogram, echo = TRUE}

```

* Calculate and report the mean and median total number of steps taken per day

```{r report_mean_and_median, echo = TRUE}
aggregate(activity$steps, list(Date = activity$date), mean)
aggregate(activity$steps, list(Date = activity$date), median)
```

* What is the average daily activity pattern?

* Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

* Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

* Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

* Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

* Create a new dataset that is equal to the original dataset but with the missing data filled in.

* Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

* Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

* Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

* Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

## Submitting the Assignment

To submit the assignment:

* Commit the your completed PA1_template.Rmd file to the master branch of your git repository (you should already be on the master branch unless you created new ones)

* Commit your PA1_template.md and PA1_template.html files produced by processing your R markdown file with knit2html() function in R (from the knitr package) by running the function from the console.

+ If your document has figures included (it should) then they should have been placed in the figure/ directory by default (unless you overrided the default). Add and commit the figure/ directory to yoru git repository so that the figures appear in the markdown file when it displays on github.

* Push your master branch to GitHub.

+ Submit the URL to your GitHub repository for this assignment on the course web site.

+ In addition to submitting the URL for your GitHub repository, you will need to submit the 40 character SHA-1 hash (as string of numbers from 0-9 and letters from a-f) that identifies the repository commit that contains the version of the files you want to submit. You can do this in GitHub by doing the following

```{r final_information, echo = TRUE}
sessionInfo()
```